---
title: Will the Result be Different if Everyone in Canada Vote for 2019 Canadian Federal Election?  -- Option b
author: "pseudonym"
date: "12/9/2020"
output: word_document
Keywords: Post-stratification, 2019 Canadian Election Survey, Canada Census, logistic regression
Abstract: The 2019 Canadian Federal Election has finished with the winner of Liberal Party. This result is derived from all the eligible Canadian Citizens’ votes. This paper will use post-stratification method and logistic models to predict the election result assuming the whole Canada population can vote. Will the winner be different?  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(visdat)
library(skimr)
```


```{r insert 2019 CES Survey Data}
# install cesR package
devtools::install_github("hodgettsp/cesR")
# load cesR package
library(cesR)
#Get the data we want which is decon
cesR::get_decon()
summary(decon)
```

## 1.Introduction
As we all know, in Canada, only Canadian citizens over 18 years old have the eligibility to vote during federal election. The winning political party was Liberal Party for 2019 and the Prime Minister is Justin Trudeau. However, there are also many other people living in Canada over 18 years old that cannot vote during election such as international students, people with a work permit, permanent citizens and refugees etc. A question comes up: what if all the people in Canada have the right to vote for their supporting political party? Will the result be different? Is this a better or worse voting patter to use for Canada’s future election? Based on this motivation and interest, I collected the survey dataset for 2019 CES and xxx Canada Census dataset to do some analysis to see whether the result would be different.
After doing a post-stratification on the xxx Canadian Census dataset and applying it to a survey dataset model for 2019 CES, we found that the result is different from the real life. I first created a proper model based on 2019 CES survey dataset including several variables of interest which may affect people’s voting choice. This model can be used to predict the voting intention according to several personal information aspects. Then, I applied the xxx Census dataset which is a post-stratification dataset onto this model, I got a different result based on the situation that all the people in Canada can vote. 
This result delicates that there should be restrictions set on people who can vote since for example, people under 18 do not have complete and mature personality and political sense. Involving the whole population will lead to a biased result. In this paper, I will clearly introduce my analyzing procedure and relative data cleaning process. Also, there are discussions about what each step means and how this result reflects towards Canada’s election pattern. 


## 2.Data 
## 2.1 Survey Data Description
pending

## 2.2 Census Data Description
pending

## 2.3 Data Cleaning 
pending

## 3.Model Construction and Diagnostics
## 3.1 Building model Based on Variables of Interest
pending
## 3.2 Diagnositics 
pending

## 4.Applying the Model to Post-Stratification Data
pending

## 5. Discussion and Result
pending

## 6. Further Thinking and Weaknesses
pending