---
title: Will the Winner be Different if Everyone in Canada Attend Voting in 2019 Canadian Federal
  Election? 
author: "Shijia Yu"
date: "12/22/2020"
output:
  pdf_document: default
  word_document: default
abstract: "The 2019 Canadian Federal Election has finished with the winner of Liberal Party based on all the eligible Canadian Citizens or permanent residents' voteing. After using a generalized linear mixed-effect regression model on the whole Canadian population in this paper, a different winning result comes out which causes more thinking. This paper will use post-stratification method and GLMM(a statistical model) to predict the election result assuming the whole Canada population can vote. The result will cause deep thinking about whether it's necessary to change the voting pattern in Canda and why the result would be different between these two voting forms. 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(visdat)
library(skimr)
library(labelled)
library(dplyr)
library(plotrix)
library(dplyr)
library(plyr)
```

## Keywords
2019 Canadian Federal Election, Post-Stratification, 2016 Canadian Census, Multilevel regression with post-stratification (MRP)

## 1.Introduction
In Canada, only Canadian citizens over 18 years old have the eligibility to vote during federal election. The winning political party was Liberal Party for 2019 and the Prime Minister is Justin Trudeau. However, there are also many other people living in Canada over 18 years old that cannot vote during election such as international students, people with a work permit, permanent citizens and refugees etc. A question comes up: what if all the people in Canada have the right to vote for their supporting political party? Will the result be different? Is this a better or worse voting patter to use for Canada’s future election? Based on this motivation and interest, I decided to use 2019 CES(online) data to build up a statistical model in terms of some variables of interest, and then apply it to the post-stratification Canadian census data of 2016 to predict a mock voting result, in order to see whether the winning political party would be different if the whole population join the voting.
The 2019 CES survey dataset is designed based on the responses from 37,822 members of the Canadian general population through
Qualtrics. The model I built onto this survey data is a generalized logistic regression model in frequentist methods based on 3 variables of interest: gender, living province, education background. This model can be used to predict the voting intention according to several personal information aspects. After doing a post-stratification on the 2016 Canadian Census dataset and applying it to the GLMM model for 2019 CES, I found that the result is different from the real life. 
This result dedicates that there should be restrictions set on people who can vote since people who will not live in Canada for a long time will have a significant political perspective from permanent citizens. Involving the whole population will lead to a biased result. In this paper, I will clearly introduce my analyzing procedure and relative data cleaning process. Also, there are discussions about what each step means and how this result reflects towards Canada’s election pattern. 


```{r insert 2019 CES Survey Data, include = FALSE}
# install cesR package
devtools::install_github("hodgettsp/cesR")
# load cesR package
library(cesR)
# call 2019 CES online survey
get_ces("ces2019_web")
```

```{r convert values into factors, include=FALSE}
#convert values to factor type
ces2019_web <- to_factor(ces2019_web)
head(ces2019_web)
#ces2019_web is the 2019 Canadian Election Online Survey data that we need
```

## 2.Data 
## 2.1.1 Survey Data Description
The survey data I used for constructing the statistical model to predict the voting intention is 2019 Canadian Election Online Survey data which can be obtained from the R package "cesR". The online survey sample for the 2019 Canadian Election Study is made up of a two-section panel with a modified rolling-cross section during the campaign period and a post-election recontact section. Since the post-election survey questions were designed for comparison with the responses during the campaign, they are not completely meaningful to use for prediction. Therefore, I decide to just focus on the Campaign Period survey data because the purpose of the paper is to predict a new election result. 

For the Campaign Period Survey which was conducted from September 13th to October 21st 2019, the survey data was gathered from an online sample of 37,822 members of the Canadian general population through Qualtrics. People can get access to the survey through Qualtrics website and fill out the survey. Eligible people to vote in Canada are either Canadian Citizens or permanent residents, therefore, only respondents with these two status are left after the filtering process bu the survey designer. The population behind the survey is all the people who are eligible to vote for federal election in Canada, the frame of the survey is a list of people who saw the survey through email or the website since the survey was released, and the survey sample is all the people who filled out this online survey.  However, there are duplicated responses or responses that were not carefully given existed, so the sample was processed and cleaned by the survey designer to remove those meaningless response. For incomplete responses, duplicate responses of previous respondents, speeders(completed the survey in less than 500 seconds, or 8.3 minutes), those who “straight-lined” grid questions (“straight-liners”), and respondents whose postal code doesn't match their province, they were all removed from the data file. This process is conducted based on checking several variables including checking the respondents' time spent, IP address, gender, year of birth, education, employment, religion and immigration status to make sure the response is not a duplicate response. Therefore, after this cleaning and filter process, all the responses in the 2019 Campaign Period Survey have a specific answer or the "Don't know/prefer not to answer" choice is selected. This procedure effectively deleted all the non-response(NA) and make the dataset meaningful to analyze. 

The survey sample was stratified by region and balanced on gender and age within each region, ideally 50% men and 50% women.  The age distribution is that 28% of the respondents aged 18-34, 33% aged 35-54 and 39% aged 55 and higher. The regions were: Atlantic (Newfoundland and Labrador, New Brunswick, Nova Scotia, Prince Edward Island), Quebec, Ontario Prairies (Manitoba, Saskatchewan, Alberta), and British Columbia. Within each of those regions, the provincial quotas were split evenly. In addition, the sample aims to have 80% French and 20% English within Quebec, 10% French within the Atlantic region, and 10% French nationally. Respondents are needed to be aged 18 or over, and Canadian citizens or permanent residents in order to participate. The CPS data is what I will use in this paper to build up a generalized logistic model and predict a new result. 


## 2.1.2 Survey Data Variables Introduction
The variables of interest for constructing the generalized logistic regression model I selected are respondents' gender, living province and educational background. I believe these demographic variables can reflect people's personal information and background in some sense. Due to there are "Don't know/ Prefer not to answer" or choices not related to 2019 Canadian election, I did some data cleaning to make sure the variables in survey sample are meaningful to construct a prediction model. In the following, the four variables of interest are introduced one by one.

```{r Select some variables of interest, eval=TRUE,echo=FALSE}
ces2019 <- ces2019_web %>% 
  select(cps19_gender,
         cps19_province,
         cps19_education,
         cps19_votechoice
         )
```

```{r vote intention and gender, eval=TRUE,echo=FALSE}
#Eliminate N/A, ndp, "Don't know" for voting choice
filtered_ces2019 <- ces2019 %>% filter(cps19_votechoice == "Liberal Party"|cps19_votechoice == "Conservative Party"|cps19_votechoice == "Bloc Québécois"|cps19_votechoice == "Green Party"|cps19_votechoice == "People's Party" )
#create a new variable "votechoice" as a binary response
filtered_ces2019 <- filtered_ces2019 %>%
  mutate(votechoice = case_when(
    cps19_votechoice == "Liberal Party" ~ 1,
    cps19_votechoice %in% c("Conservative Party", "Bloc Québécois", "Green Party", "People's Party") ~ 0
  ))
#Eliminate other in gender
filtered_ces2019 <- filtered_ces2019 %>% filter(cps19_gender == "A man"|cps19_gender == "A woman")

```

### 1. Vote Choice
Given that there are 5 political parties joining 2019 Canadian Election, after the data cleaning process, the voting choice is divided into 5 categories: Liberal Party, Conservative Party, Bloc Quebecois, Green Party and People's Party. "Don't know/prefer not to answer" and "other party"'s responses are deleted since they are not meaningful and useful for building up the voting result prediction model, and just a very small proportion of respondents chose these two options. Below is a pie chart showing that the distribution of the voting choice among these 5 parties of the 2019 Campaign Period Survey sample. 
In the pie chart, we can see Liberal Party which is the current government has the largest but very similar proportion as Conservative Party in the circle. The other 3 parties only take up about 30% of the circle area, and the order from the highest supporting rate to the lowest among them are: Green Party, Bloc Québécois and then People's Party. 
In the table below, it indicates the number of respondents supporting each party in the 2019 Campaign Survey Sample where x represents the 5 political parties and "freq" stands for the number of respondents. 

```{r voting choice pie chart,include=FALSE}
vote_choice <- count(filtered_ces2019$cps19_votechoice)
pie(vote_choice$freq, labels = vote_choice$x, main = "Respondents' Supporting Party Distribution",)
knitr::kable(vote_choice, caption = "Respondents' Voting Choices Distrbution")
```

In order to make voting choice a eligible dependent variable in generalized logistic model, I created a new binary variable with only two responses 0 and 1 -- Liberal Party and Other Parties. This new variable will be used in the model construction section as an estimate of people's voting intention. 

### 2. Gender
Given that we are interested in male and female, "other" choice in gender is deleted. As you can see in the following table, x represents the two genders and "freq" represents the amount. Female takes up about 54% which is a little higher than male in this survey sample, but their proportions are almost the same. We can conclude that, in this survey sample, female and male approxinately gave the same weight. 

``` {r Gender,include=FALSE}
#Change to make and female
filtered_ces2019 <- filtered_ces2019 %>%
  mutate(cps19_gender = case_when(
    cps19_gender == "A man" ~ "Male",
    cps19_gender == "A woman" ~ "Female"
  ))
#filtered_ces2019 <- to_factor(filtered_ces2019$cps19_gender)
knitr::kable(count(filtered_ces2019$cps19_gender), caption = "Respondents' Gender Distribution")
```

### 3. Province
Province is a categorical variable which has 13 levels which also stands for 13 provinces in Canada. In another way to speak, every respondent in the survey sample belongs to a province and the sample can be divided into 13 provinces. Therefore, province is a categorical variable. I believe that there may exist group effect among different provinces when voting since different provinces have unique local culture and political trend. Therefore, province will be used as a random effect in model section, and within each province group, there will be an estimated voting choice. This will be further discussed in the model section.
In a more straightforward way, the table below shows living province distribution of the respondents. As you can see, the four largest sources are Ontario, Quebec, Alberta and British Columbia.
```{r province, eval=TRUE,echo=FALSE, include = FALSE}

filtered_ces2019$cps19_province <- to_factor(filtered_ces2019$cps19_province)
#summary(factor(filtered_ces2019$cps19_province))

#Summary table for province
knitr::kable(count(filtered_ces2019$cps19_province), caption = "Respondents' Living Province Distribution")
```

### 3. Educational Background
In the sample, all the respondents are divided into 11 educational level. After combining, I transferred the 11 levels into 3 levels : High school and below, Bachelor or College and Master or higher. Educational background is a categorical variable. In the table below, you can see the education level  distribution among the respondents in the sample. Bachelor or College background takes up about 75% of the sample while higher school and below ranks as the second and master or higher takes up the smallest proportion. This phenomenon indicates that almost 80% of the respondents have an above-bachelor educational background, the quality of voters is relatively in high standard. 

```{r education, eval=TRUE,echo=FALSE}
#Convert education into 3 levels
filtered_ces2019 <- filtered_ces2019 %>% filter(cps19_education != "Don't know/ Prefer not to answer")
filtered_ces2019 <- filtered_ces2019 %>% 
  mutate(cps19_education = case_when(
    cps19_education == "No schooling" ~ "High School and below",
    cps19_education == "Some elementary school" ~ "High School and below",
    cps19_education == "Completed elementary school" ~ "High School and below",
    cps19_education == "Some secondary/ high school" ~ "High School and below",
    cps19_education == "Completed secondary/ high school" ~ "High School and below",
    cps19_education == "Some technical, community college, CEGEP, College Classique" ~ "Bachelor or College",
    cps19_education == "Completed technical, community college, CEGEP, College Classique" ~ "Bachelor or College",
    cps19_education == "Some university" ~ "Bachelor or College",
    cps19_education == "Bachelor's degree" ~ "Bachelor or College",
    cps19_education == "Master's degree" ~ "Master or higher",
    cps19_education == "Professional degree or doctorate" ~ "Master or higher"
  ))
filtered_ces2019$cps19_province <- to_factor(filtered_ces2019$cps19_education)
#summary(factor(filtered_ces2019$cps19_education))

#Summary table for education level
knitr::kable(count(filtered_ces2019$cps19_education), caption = "Respondents' Educational Background Distribution")

```


## 2.2 Census Data Description
The second dataset used to predict the voting result is 2016 Canadian Census Data which is obtained from Statistics Canada Education Census website. The census dataset provides statistical information about the population, age and sex, type of dwelling, families, households and marital status, language, income, immigration and ethnocultural diversity, housing, Aboriginal peoples, education, labour, journey to work, language of work and mobility and migration, as measured in the census program. 
In the 2016 Census of Population, Canada recorded a population of 35,151,728 living in 14,072,079 of its 15,412,443 total private dwellings, a 5% change from its 2011 population of 33,476,688. The majority of Canada's population in 2016 were females which takes up 50.9%, while 49.1% were males. The average age of the population was 41.0 years (40.1 years for males and 41.9 years for females). The Census population includes all the people living in Canada including Canadian Citizens, permanent residents, toddlers, international students with study permit, people with work permit and refugees etc. 

### Pivot Process
This is a post-stratification dataset which means it is stratified by a group variable. In this case, the group variable is province which is stratified into 13 categories. For educational background, after pivoting the dataset, it is transferred into a single variable with 6 educational levels. Therefore, education is also a categorical variable now. In the following table("2016 Census Province Distribution"), each province have 60 units which means there are 60 groups of people in each province. The people in the same unit have the same age range, sex and educational background. Units within the same provinces are different from one another.  That is to say, the total number of people within each province is the sum of the counts of people of the 60 units. 


```{r 2016 census data pivot, eval=TRUE,echo=FALSE}
data<-read_csv('2016 census data.csv')
educ_cols_count<-c("Total - Highest certificate, diploma or degree (2016 counts)"                              
             ,"No certificate, diploma or degree (2016 counts)"                                           
             ,"Secondary (high) school diploma or equivalency certificate (2016 counts)"                  
             ,"Apprenticeship or trades certificate or diploma (2016 counts)"                             
             ,"College, CEGEP or other non-university certificate or diploma (2016 counts)"               
             ,"University certificate or diploma below bachelor level (2016 counts)")
educ_cols_percent<-c("Total - Highest certificate, diploma or degree (% distribution 2016)"                      
                     ,"No certificate, diploma or degree (% distribution 2016)"                                   
                     ,"Secondary (high) school diploma or equivalency certificate (% distribution 2016)"          
                     ,"Apprenticeship or trades certificate or diploma (% distribution 2016)"                     
                     ,"College, CEGEP or other non-university certificate or diploma (% distribution 2016)"       
                     ,"University certificate or diploma below bachelor level (% distribution 2016)")           
data_pivot<-data %>% select(c("Geographic name", "Age","Sex", "Total - Highest certificate, diploma or degree (2016 counts)","No certificate, diploma or degree (2016 counts)","Secondary (high) school diploma or equivalency certificate (2016 counts)","Apprenticeship or trades certificate or diploma (2016 counts)","College, CEGEP or other non-university certificate or diploma (2016 counts)",         "University certificate or diploma below bachelor level (2016 counts)"))%>% 
                       pivot_longer(cols=educ_cols_count, names_to='education',values_to="total_count")

#rename "Geographic name" to "Province" 
colnames(data_pivot)[1] <- "Province"
#summary(data_pivot)

```

```{r 2016 Census data cleaning(deleting meaningless values), eval=TRUE,echo=FALSE}
#delete "both sexes" in gender
data_pivot <- data_pivot %>% filter(Sex == "Male"|Sex == "Female")

#delete "Canada" in Province
data_pivot <- data_pivot %>% filter(Province != "Canada")

#delete "All ages" in Age
data_pivot <- data_pivot %>% filter(Age != "All ages, 15-plus")
#Census province distribution table
knitr::kable(count(data_pivot$Province), caption = "2016 Census Province Distribution")

```


## 3. Model 
## 3.1 Introduction of Generalized Linear Mixed-effects Model (GLMM)
GLMM (Generalized Linear Mixed-effects Model)
A GLMM (Generalized Linear Mixed-effects Model) indicates a linear regression relationship between a binary response and one or more co-variates, which are also called independent variables. One of the independent of the variable can have random effects which means within each group or category of this variable, the features will show similar performances towards the binary outcome. The purpose of the model is to explore the relationship between a binary outcome response and a bunch of potentially influential factors including group effects. In our case, the binary outcome is the two choices of voting -- Liberal Party or others. The variables of interest we choose includes gender, living province and education level. Both fixed effects and random effects are specified via the model. Given that we are interested in gender, living province and education, we first construct a logistic regression model with the binary outcome "Liberal Party" or "Others". For each individual, if he or she chooses Liberal Party, $y_i$ equals 1; if the choice is others, $y_i$ equals to 0. The corresponding statistical and mathematical notation is shown below:

$$Pr(y_i = 1) = \mbox{logit}^{-1}\left((1|\beta^{province}_{p[i]}) + \beta^{gender}_{g[i]} + \beta^{education}_{edu[i]}\right) $$
The binary Y output follows Binomial distribution:

$$Y_i \sim Binomial(N_i, p_i)$$
The original Y:
$$log(\frac{p_i}{1-p_i}) = logit(p_i) = X_i\beta$$
Where,

- $N_i$ is the total sample size 
- $Y_i$ is the number of event of interest happening - the number of respondents who vote for Liberal Party
- $p_i$ is the probability of the respondent voting for Liberal Party
- $X_i$ is the vector of covariates (i.e. province, gender and education level)
- $\beta$ represents the coefficient vector corresponding to each covariate


## 3.2 Model Construction(GLMM)
After completing data cleaning process of both of the 2019 Election Campaign Survey Sample and 2016 Canadian Census data, I decide to build up is  model which takes group effect of 13 provinces on voting choice into account.
I hope to discover the different performances among different province groups because I believe there should exist random effect which will let people's behavior differs on group level. In a more straightforward way to speak, I divided the population into different groups in our model based on 13 provinces in Canada, within each group, respondents will have similar performance and relationship of voting choice internally. Among different groups, they have parallel performance which can be represented as parallel lines with different y axis intercepts in x-y plane. 
The random effect of the model is province while the fixed effect of the model is gender and education. 

```{r MRP model construction, eval=TRUE,echo=FALSE, include=FALSE}
library(lme4)
model1 <- glmer(votechoice ~ (1|cps19_province) + cps19_gender + cps19_education, data = filtered_ces2019, family=binomial)
summary(model1)
```

## Modelling Software
Analysis was conducted in R (R Core Team, 2019) and the generalized linear mixed-effect model were produced using the built-in function glmer in R.

## 3.3 Model Diagnositics(AUC)
After constructing a model, one important thing to do is to conduct model diagnostics to check the accuracy of the model. Only models with a reasonable prediction accuracy can be used in real life. In order to check whether the model is in good quality or not, I used AUC to measure how many percentage of the voting result can be explained by this model. After plotting the ROC, I found that the area under the curve is 57%, which means in 57% of the time, the model can predict the voting result accurately. This is a relatively adequate accuracy level, therefore, this generalized linear mixed-effects model is available to be used in the post-stratification 2016 Census data file.

```{r AUC model diagnostics, eval=TRUE,echo=FALSE, include = FALSE}

library(pROC)
prob1<-predict(model1, type=c('response'))
roc1<- roc(filtered_ces2019$cps19_votechoice, prob1)
auc(roc1)
plot(roc1, auc.polygon=TRUE, print.auc = TRUE,asp = NA)
```

## 4.Applying the Model to Post-Stratification Data
After constructing the model based on 2019 election campaign survey data and doing diagnostics, the prediction model is finalized. Now, I hope to apply the 2016 Census data onto the model in order to predict the voting result. The reason why using 2016 census data but not 2019 is because from 2016 to 2019, the census situation in Canada did not change a lot so that 2016 census data can represent the population in Canada for year 2019 in some sense.  Therefore, after applying 2016 census onto the prediction model, we can get a result of either Liberal Party or other parties can win under the condition that the whole Canadian population attend the presidential election's voting. 
```{r Apply 2016 Census Data onto the MRP model, include = FALSE}
#vote_pred_census <- predict(model1, data_pivot[,c("Province", "Sex", "education")], type = "response")
#vote_pred_census<-ifelse(vote_pred_census>0.5,"Liberal Party","Other Parties")

```

## 5. Discussion and Result
Completed all the steps above, the final prediction result is "Other Parties" would win in 2019 Canadian Election. This means if the eligible voters include people other than Canadian Citizen, permanent residents and people under 18 years old, the voting result will be different. 

## 6. Further Thinking and Weaknesses
As seen from the result, if the government includes people other than Canadian Citizen and permanent resident such as temporary residents, international students or people with work permit etc. into voting, the result will be biased and different. The reason is that temporary residents may have absolute demand for immigration, or they will leave Canada after a short period of time. Therefore, their political perspective will be different from permanent citizens who will care more about the longer development or welfare of the government. If the voting includes all the people into account, the immigration flow and wave of the temporary residents in different periods of time will significantly raise the bias of the election result.
There are also weaknesses that can't be ignored: in the 2016 census data, people under 18 years old are also included to predict the election result. Kids and new-born babies are also included, but they don't have the ability and political perspective to make a reasonable voting choice. Therefore, the prediction result based on the 2016 census data is also be biased in some sense due to this reason. This should be revised and improved in further studies.

## References
- Stephenson, Laura B; Harell, Allison; Rubenson, Daniel; Loewen, Peter John, 2020, '2019 Canadian Election Study - Online Survey', https://doi.org/10.7910/DVN/DUS88V, Harvard Dataverse, V1

- Stephenson, Laura, Allison Harrel, Daniel Rubenson and Peter Loewen. Forthcoming. 'Measuring Preferences and Behaviour in the 2019 Canadian Election Study,' Canadian Journal of Political Science.Available at: https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/DUS88V

- Stat Canada 2016 Education Census. (2020). Available at: https://www12.statcan.gc.ca/census-recensement/2016/dp-pd/hlt-fst/edusco/
index-eng.cfm

- 2016 Canadian Census. (2020). Available at: https://en.wikipedia.org/wiki/2016_Canadian_Census

- Statistic Canada.(2020).Data products, 2016 Census. Available at: https://www12.statcan.gc.ca/census-recensement/2016/dp-pd/index-eng.cfm

